Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging

### 2022-12-07 script/manuscript_MEE.Rmd
stylinadded results for temporal, need to checkin with paul and jane to have them clean it up and probably drop the tau for a simpler test, splitting appendixes drop em in

  **KEYWORDS**    
   DNA barcodes, Plantae, metagenomics, molecular barcoding, species distribution modelling, Angiosperms353, bombus, queen foraging.
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: true
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

Pollen genomic DNA was extracted from corbiculae using a CTAB based protocol modified from Lahlamgiahi et al. and Guertler et al. (2014, 2014). A SDS extraction buffer (350µL , 100mM Tris-HCl, 50 mM EDTA, 50 mM NaCl, 10% SDS v/v., pH 7.5) was added followed by vortexing to allow dissolution of corbiculae. Pollen grains were then macerated with Kontes Pellet Pestles, and the tip of these washed with 130 µL of the SDS extraction buffer, samples were then incubated for 1 hour at 30°C. This was followed by the addition of 10% CTAB solution (450ul, of 20 mM Tris-Cl pH. 8.0, 1.4 M NaCl, 10 mM EDTA pH 7.5, 10% CTAB, 5% PVP, ~85% Deionized water) and RNAse (10 uL of 10 mg/mL) and samples were incubated for 40 minutes at 37°^C^, on heat block (Multi-Blok, Thermo Fisher Scientific, Waltham Massachusetts) set to 40°C. After 20 minutes incubation, Proteinase K (15 µL of 20mg/ml) and  DTT (12.5 µL of 1M in water) were added, and the samples were further incubated at 60°c for 1 hour. Samples were then incubated overnight at 40°^C^. 500 µL of Phenol-Chloroform-Isoamyl alcohol (25:24:1) were added, vortexed, and centrifuged at 10,000 rpm for 10 minutes and the aqueous phase was pipetted to a 1.5 ml centrifuge tube.

@@

To develop a reference library of  pollen grains which may be present in corbiculae loads, an image reference collection of fuchsin-jelly stained (@beattie1971technique) slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@brosi2013single). Using 5 years of observational data on *Bombus* Queen Bee foraging at these studies sites (Ogilvie unpublished), as well as the Vascular Plant Checklist (@fraser2007vpc), an additional 62 voucher slides for species were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  In order to determine which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes, Divisive Hierarchical Clustering techniques were used. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (APPENDIX NO.). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner. 
  To prepare the pollen slides from corbiculae, all corbiculae loads were broken apart and rolled using dissection needlepoints to increase heterogeneity of samples. *Cerca* 0.5mm^2 of pollen was placed onto a ~4mm^2 fuchsin jelly cube (@beattie1971technique) atop a graticulated microscope slide, with 20 transects and 20 rows (400 quadrats) (EMS, Hartfield, PA). The jelly was melted, with stirring, until pollen grains were homogeneously spread across the microscope slide. Slides were sealed with Canada Balsam (Rublev Colours, Willits, CA) followed by sealing with nail polish; all samples are noted in *APPENDIX 3*.
  To identify the pollen present in corbiculae loads, light microscopy at 400x (Zeiss Axioscope A1) was used. In initial sampling in three transects, each pollen grain was identified to morphotype and counted; an additional two transects were scanned for morphotypes unique to that slide, if either transect contained an unique morphotype than all grains in that transect were also identified and counted. Subsequent to the first round of sampling, non-parametric species richness rarefaction curves (@vegans2022), and non-parametric species diversity rarefaction curves were used to assess the completeness of sampling (@inextArticle, @inextPackage). Slides not approaching the asymptote of the rarefaction curve were then re-sampled, and analysed iteratively for up to a total of seven transects *APPENDIX 2*.

@@

To estimate the duration of dates in which plant species were flowering weibull estimates of several phenologica lparameters all spatially modelled taxa were developed (@belitz2020accuracy, @pearse2017statistical). Only BIEN records which occurred in the Omernik Level 4 Ecoregions within 15km of the study area (n = 5, or conditionally 6 if enough records not be found in the nearest 5), and which were from herbarium records were included. To remove temporally irrelevant herbarium records, i.e. material collected during times which flowering is impossible at the study area due to snow cover, the SnowUS dataset (@iler2021conceptual, @tran2019cloud) from 2000-2017 was analyzed for the first three days of contiguous snow absence, and the first three days of contiguous snow cover in Fall. Herbarium records after the 3rd quantile for melt, and the 1st quantile for snow cover of these metrics were removed. Species with > 10 records had their weibull distributions generated for the date when 10% of individuals had begun flowering, when 50% were flowering, and when 90% of individuals had flowered.

@@

At the field site, of the 554 vascular plants with biotic pollination syndromes, the 493 ML ensembles accurately predicted the presence of 362 (65.3%), incorrectly predicted the presence of 64 (11.6%), incorrectly predicted 34 true presences (6.1%) as being absent, and correctly predicted the true absence of 33 (6.0%). The balanced accuracy of the ensembled models is 0.627 (Sensitivity = 0.340, Specificity 0.914).  Of the Of the 554 vascular plants with biotic pollination syndromes, the 475 LM ensembles accurately predicted the presence of 286 (51.6%), incorrectly predicted the presence of 41 (14.3%), incorrectly predicted 93 true presences (16.8%) as being absent, and correctly predicted the true absence of 55 (9.9%). The balanced accuracy of the ensembled models is 0.664 (Sensitivity = 0.573, Specificity 0.754). Of the 554 vascular plants with biotic pollination syndromes in the flora 13 (2.3%) were in the Orchid family and 41 (7.4%) are non-natives, both of which are restricted from the database, and can only reduce the number of true predicted presences by roughly 10%.

@@

no_loci <- loci %>% group_by(Sample) %>% count(name = 'no_loci')
no_reads <- loci %>% group_by(loci) %>% summarize(readsLoci = sum(n))

rm(prettyNum)
```

@@

Only 58 of these 388 species (n = 34.56897, Mdn = 31) were able to be compared to ground truth data from the long term (1974–2012) dataset. Of these species relatively high accord was observed between the long-term ground truthed dataset, and the modelled species.

@@

Although our temporal results were lackluster, we note that our study area has an incredibly brief growing period. and we suspect these temporal results would be useful in sub-tropical and tropical ecosystems. FURTHER, the sites used for ground truthing the temporal flowering periods were not randomly selected across the study area, and cannot be used to make inference to the population across the entirety of the study area as we did here. Regardless both show good agreement on flower onset, peak flowering, and moderate agreement with flowering cessation. The disagreement in flowering cessation is perhaps due to more microclimates which retain water, rather than microclimates which allow the early accumulation of heat.

@@

**ACKNOWLEDGEMENTS:** Nyree Zerega for assistance obtaining herbaria loans and accessioning our collections at CHIC. Pat Herendeen for assistance with virtually all aspects of preparing pollen vouchers and the identification process. We thank the curators at the following herbaria for supplying tissue: Ben Legler at Stillinger (ID), Charles Williams at Ray J. Davis (IDS), Ernie Nelson at Rocky Mountain (RM). We thank the original collectors of these specimens: D. Knoke, L. Brummer, J. Boyd, C. Davidson, I. Gilman, M. Kirkpatrick, S. McCauley, J. Smith, K. Taylor, & C. Williams. David Giblin for sharing relevant sections of an advanced draft of FNA V. 15. Zoe Diaz-Martinez, Angela McDonnell, & Elena Loke for assistance with genomic library preparation. The Bureau of Land Management is thanked as many plant specimens were collected by R.C.B as a partner or contractor to the agency, Sarah Burnett and Lauren Price are thanked for sharing AIM data. Sanda and NEB are gratefully acknowledged for technical support and generously sharing samples.

@@

@@

```{r APPENDIX NUMBER RMBL Pollen Key, echo = F}
knitr::include_graphics('../graphics/assorted/RMBL_pollen_key.pdf')
```

@@

```{r APPENDIX MORPHOTYPE RICHNESS AND ABUNDANCE, echo= F}
knitr::include_graphics('../graphics/plots/species_richness_rarefaction.png')
knitr::include_graphics('../graphics/plots/SppAbundance.png')
```

>>>>

output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: true
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

Pollen genomic DNA was extracted from corbiculae using a CTAB based protocol modified from Lahlamgiahi et al. and Guertler et al. (2014, 2014). A SDS extraction buffer (350µL , 100mM Tris-HCl, 50 mM EDTA, 50 mM NaCl, 10% SDS v/v., pH 7.5) was added followed by vortexing to allow dissolution of corbiculae. Pollen grains were then macerated with Kontes Pellet Pestles, and the tip of these washed with 130 µL of the SDS extraction buffer, samples were then incubated for 1 hour at 30°C. This was followed by the addition of 10% CTAB solution (450ul, of 20 mM Tris-Cl pH. 8.0, 1.4 M NaCl, 10 mM EDTA pH 7.5, 10% CTAB, 5% PVP, ~85% Deionized water) and RNAse (10 uL of 10 mg/mL) and samples were incubated for 40 minutes at 37°C, on heat block (Multi-Blok, Thermo Fisher Scientific, Waltham Massachusetts) set to 40°C. After 20 minutes incubation, Proteinase K (15 µL of 20mg/ml) and  DTT (12.5 µL of 1M in water) were added, and the samples were further incubated at 60°C for 1 hour. Samples were then incubated overnight at 40°C. 500 µL of Phenol-Chloroform-Isoamyl alcohol (25:24:1) were added, vortexed, and centrifuged at 10,000 rpm for 10 minutes and the aqueous phase was pipetted to a 1.5 ml centrifuge tube.

@@

To develop a reference library of  pollen grains which may be present in corbiculae loads, an image reference collection of fuchsin-jelly stained (@beattie1971technique) slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@brosi2013single). Using 5 years of observational data on *Bombus* Queen Bee foraging at these studies sites (Ogilvie unpublished), as well as the Vascular Plant Checklist (@fraser2007vpc), an additional 62 voucher slides for species were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  In order to determine which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes, Divisive Hierarchical Clustering techniques were used. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (APPENDIX NO.). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner. 
  To prepare the pollen slides from corbiculae, all corbiculae loads were broken apart and rolled using dissection needlepoints to increase heterogeneity of samples. *Cerca* 0.5mm^2 of pollen was placed onto a ~4mm^2 fuchsin jelly cube (@beattie1971technique) atop a graticulated microscope slide, with 20 transects and 20 rows (400 quadrants) (EMS, Hartfield, PA). The jelly was melted, with stirring, until pollen grains were homogeneously spread across the microscope slide. Slides were sealed with Canada Balsam (Rublev Colours, Willits, CA) followed by sealing with nail polish; all samples are noted in *APPENDIX 3*.
  To identify the pollen present in corbiculae loads, light microscopy at 400x (Zeiss Axioscope A1) was used. In initial sampling in three transects, each pollen grain was identified to morphotype and counted; an additional two transects were scanned for morphotypes unique to that slide, if either transect contained an unique morphotype than all grains in that transect were also identified and counted. Subsequent to the first round of sampling, non-parametric species richness rarefaction curves (@vegans2022), and non-parametric species diversity rarefaction curves were used to assess the completeness of sampling (@inextArticle, @inextPackage). Slides not approaching the asymptote of the rarefaction curve were then re-sampled, and analysed iteratively for up to a total of seven transects *APPENDIX 2*.

@@

To estimate the duration of dates in which plant species were flowering weibull estimates of several phenological parameters all spatially modelled taxa were developed (@belitz2020accuracy, @pearse2017statistical). Only BIEN records which occurred in the Omernik Level 4 Ecoregions within 15km of the study area (n = 5, or conditionally 6 if enough records not be found in the nearest 5), and which were from herbarium records were included. To remove temporally irrelevant herbarium records, i.e. material collected during times which flowering is impossible at the study area due to snow cover, the SnowUS dataset (@iler2021conceptual, @tran2019cloud) from 2000-2017 was analyzed for the first three days of contiguous snow absence, and the first three days of contiguous snow cover in Fall. Herbarium records after the 3rd quantile for melt, and the 1st quantile for snow cover of these metrics were removed. Species with > 10 records had their weibull distributions generated for the date when 10% of individuals had begun flowering, when 50% were flowering, and when 90% of individuals had flowered.

@@

At the field site, of the 554 vascular plants with biotic pollination syndromes, the 493 ML ensembles accurately predicted the presence of 362 (65.3%), incorrectly predicted the presence of 64 (11.6%), incorrectly predicted 34 true presences (6.1%) as being absent, and correctly predicted the true absence of 33 (6.0%). The balanced accuracy of the ensembled models is 0.627 (Sensitivity = 0.340, Specificity 0.914). Of the 554 vascular plants with biotic pollination syndromes, the 475 LM ensembles accurately predicted the presence of 286 (51.6%), incorrectly predicted the presence of 41 (14.3%), incorrectly predicted 93 true presences (16.8%) as being absent, and correctly predicted the true absence of 55 (9.9%). The balanced accuracy of the ensembled models is 0.664 (Sensitivity = 0.573, Specificity 0.754). Of the 554 vascular plants with biotic pollination syndromes in the flora 13 (2.3%) were in the Orchid family and 41 (7.4%) are non-natives, both of which are restricted from the database, and can only reduce the number of true predicted presences by roughly 10%.

@@

no_loci <- loci %>% group_by(Sample) %>% count(name = 'no_loci')
no_reads <- loci %>% group_by(loci) %>% summarize(readsLoci = sum(n))
```

@@

Only 58 of these 388 species (n = 34.56897, Mdn = 31) were able to be compared to plot based observational data from the long term (1974–2012) dataset. Of these species relatively high accord was observed between the long-term ground truthed dataset, and the modelled species. There was very strong evidence that the weibull estimates were positively associated with the observed onset (r^2^ = 0.72, p < 0.0001, tau = 0.61) and peak (r^2^ = 0.70, p < 0.0001, tau = 0.65) of flowering, and that the number of herbarium samples had a moderate effect on the estimates (p = 0.004 and p = 0.034 respectively). There was very strong evidence that the weibull estimates had a positive association with the observed cessation of flowering (r^2^ =  0.4339,  p < 0.0001, tau = 0.489), however their was no evidence that sample size had an effect (p = 0.349). There was moderate evidence that the weibull estimates, with an effect of sample size, had a weak positive association with the observed duration of flowering (p = 0.0401, r^2^ = 0.07, tau = 0.17).

@@

** Mention in here that the general effect of sample size on predictions means that larger samples sizes are required for this to work!!!! **
Although our temporal results were lackluster, we note that our study area has an incredibly brief growing period. and we suspect these temporal results would be useful in sub-tropical and tropical ecosystems. FURTHER, the sites used for ground truthing the temporal flowering periods were not randomly selected across the study area, and cannot be used to make inference to the population across the entirety of the study area as we did here. Regardless both show good agreement on flower onset, peak flowering, and moderate agreement with flowering cessation. The disagreement in flowering cessation is perhaps due to more microclimates which retain water, rather than microclimates which allow the early accumulation of heat.

@@

**ACKNOWLEDGMENTS:** Nyree Zerega for assistance obtaining herbaria loans and accessioning our collections at CHIC. Pat Herendeen for assistance with virtually all aspects of preparing pollen vouchers and the identification process. Hilary Noble, Zoe Diaz-Martinez, Angela McDonnell, & Elena Loke for assistance with genomic library preparation. We thank the curators at the following herbaria for supplying tissue: Ben Legler at Stillinger (ID), Charles (Rick) Williams at Ray J. Davis (IDS), (B)Ernie Nelson at Rocky Mountain (RM). We thank the original collectors of these specimens: D. Knoke, L. Brummer, J. Boyd, C. Davidson, I. Gilman, M. Kirkpatrick, S. McCauley, J. Smith, K. Taylor, & C. Williams. David Giblin for sharing relevant sections of an advanced draft of FNA V. 15. The Bureau of Land Management is thanked as many plant specimens were collected by R.C.B as a partner or contractor to the agency, Sarah Burnett and Lauren Price are thanked for sharing AIM data. Sanda and NEB are gratefully acknowledged for technical support and generously sharing samples.

@@

\newpage

@@

\newpage

```{r APPENDIX NUMBER RMBL Pollen Key 1, echo = F}
knitr::include_graphics('../graphics/assorted/RMBL_pollen_key-1.pdf')
```

\newpage

```{r APPENDIX NUMBER RMBL Pollen Key 2, echo = F}
knitr::include_graphics('../graphics/assorted/RMBL_pollen_key-2.pdf')
```

\newpage

@@

\newpage

```{r APPENDIX MORPHOTYPE RICHNESS AND ABUNDANCE, echo= F}
knitr::include_graphics('../graphics/plots/species_richness_rarefaction.png')
```

\newpage

```{r, echo = F, out.width = '25%'}
knitr::include_graphics('../graphics/plots/SppAbundance.png')
```

\newpage

```{r Voucher Reference Specimen tables 1, echo = F}
knitr::include_graphics('../graphics/tables/specimen_tables-1.pdf')
```

\newpage

```{r Voucher Reference Specimen tables 2, echo = F}
knitr::include_graphics('../graphics/tables/specimen_tables-2.pdf')
```

\newpage
